Identifying optimal baselines for stable isotope studies of stream food webs

### 2022-05-11 docs/ms.Rmd
add .bib and update ms

---
title: "Identifying optimal baselines for stable isotope studies of stream food webs"
author: 
  - "Nathan Barrus^1^, Bryan M. Maitland^1,2^, Frank J, Rahel^1^"
  - "^1^University of Wyoming" 
  - "^2^UW-Madison and WDNR"
date: "`r format(Sys.time(), '%Y-%B-%d')`"
output: 
  # word_document: default
  # pdf_document:
  #   fig_caption: yes
  #   keep_tex: yes
  #   latex_engine: pdflatex
  bookdown::html_document2:
    number_sections: false
fontsize: 11pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{url}
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
# bibliography: baseline-isotopes.bib
# csl: freshwater-biology.csl
---

@@

```{css, echo=FALSE}
h1, h4 {
  text-align: center;
}

@@

Stable isotope analysis (SIA) is a commonly-used tool for making inferences about the trophic structure and flow of materials in food webs. Stable isotope ratios—typically of carbon (13C/12C, d13C) or nitrogen (15N/14N, d15N)—provide information on the materials assimilated by animals because of predictable changes in isotopic composition that occur in body tissues during elemental incorporation of diet items . With each trophic transfer, carbon isotope ratios change little which makes them useful for tracking energy pathways from basal resources and identifying the ultimate carbon source supporting consumers. In contrast, nitrogen isotope ratios change markedly and predictably with each trophic transfer, allowing for inference into species’ trophic positions in communities, as well quantification of whole-community food chain length (Post, 2002; Mccutchan et al., 2016; Pinzone et al., 2017).

@@

When possible, we used individual macroinvertebrates for SIA but for small taxa we pooled individuals to ensure enough material for SIA. Muscle fillets were removed from the cranial dorsal portion of each fish. Samples were oven died (60 $^\circ$C, 48 hours), ground into a homogenous powder, and weighed to the nearest 0.001 mg. Animal samples (~1.0 mg) and basal resources (~2.0 mg) were weighed into 8 x 5-mm tin capsules and analyzed for d15N, d13C, %C, %N and C:N ratio at the University of Wyoming Stable Isotope Facility using a Delta Plus XP Continuous Flow Stable Isotope Ratio Mass Spectrometer (Thermo Finnigan, Bremen, Germany) coupled to a Costech Analytical 4010 elemental analyzer. Carbon (13C:12C) and nitrogen (15N:14N) isotope ratios (R) were estimated relative to their respective International Atomic Energy Agency standards (Pee Dee Belemnite limestone and atmospheric nitrogen). Working internal standards were run as controls throughout analysis to ensure accurate measurements. Isotopic values are reported in per mill delta notation \@ref(eq:ratio):

@@

For all macroinvertebrate taxa and FFGs, we used least squares linear regression to assess relationships between the baseline’s d15N position at sites along the longitudinal stream gradient.  We wished to compare our baselines’ relationships to the longitudinal gradient with those of the basal resources. So, we did separate least squares linear regressions for two of the basal resources at our sites and compared R2 values between the baselines and the resources at our sites.  The R2 value of our basal resources represent the minimum level of variation that our alternate baselines should explain in order to be used as alternatives to the traditional method (e.g. all basal resources across a season).   Baselines with significant ($P \le 0.05$) relationships with the longitudinal gradient and that had R2 values equal to or greater than the basal resources were considered suitable baselines.

@@

\begin{equation}

\end{equation}

where TPConsumer is the trophic position of each individual fish at a site, d15NConsumer is the d15N signature of each individual fish, d15NBaseline is the mean d15N signature of the baseline at each site, ∆15N is the discrimination factor for each trophic transfer, and TLBaseline is the expected trophic level of the baseline. Discrimination factor were estimated using the methods of Caut et al. (2009).

@@

\begin{equation}

\end{equation}

@@

The two basal resources, biofilm and seston, had d15N signatures that were significantly (P ≤ 0.05) correlated with the longitudinal stream gradient (PC1) (Figure 3; Table 1). All eight of the well distributed taxa and all FFG’s had d15N signatures that were significantly correlated with PC1 (Figure 3). The amount of variation in d15N signatures captured by the longitudinal gradient was similar among the two basal resources, eight sufficiently distributed taxa and the four FFGs.

>>>>

---
title: "Identifying optimal baselines for stable isotope studies of stream food webs"
author: 
  - "Nathan Barrus^1^, Bryan M. Maitland^1,2^, Frank J, Rahel^1^"
  - "^1^University of Wyoming" 
  - "^2^UW-Madison and WDNR"
date: "`r format(Sys.time(), '%Y-%B-%d')`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  bookdown::html_document2:
    number_sections: false
  # word_document: default
fontsize: 11pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{url}
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
bibliography: baseline-isotopes.bib
csl: freshwater-biology.csl
editor_options: 
  chunk_output_type: console
---

@@

```{css, echo=FALSE}
h1, h4.author, h4.date {
  text-align: center;
}

@@

Stable isotope analysis (SIA) is a commonly-used tool for making inferences about the trophic structure and flow of materials in food webs. Stable isotope ratios—typically of carbon (13C/12C, d13C) or nitrogen (15N/14N, d15N)—provide information on the materials assimilated by animals because of predictable changes in isotopic composition that occur in body tissues during elemental incorporation of diet items [@VanderZanden1997; @Perkins2014]. With each trophic transfer, carbon isotope ratios change little which makes them useful for tracking energy pathways from basal resources and identifying the ultimate carbon source supporting consumers. In contrast, nitrogen isotope ratios change markedly and predictably with each trophic transfer, allowing for inference into species’ trophic positions in communities, as well quantification of whole-community food chain length (Post, 2002; Mccutchan et al., 2016; Pinzone et al., 2017).

@@

<!-- When possible, we used individual macroinvertebrates for SIA but for small taxa we pooled individuals to ensure enough material for SIA. Muscle fillets were removed from the cranial dorsal portion of each fish. Samples were oven died (60 $^\circ$C, 48 hours), ground into a homogenous powder, and weighed to the nearest 0.001 mg. Animal samples (~1.0 mg) and basal resources (~2.0 mg) were weighed into 8 x 5-mm tin capsules and analyzed for d15N, d13C, %C, %N and C:N ratio at the University of Wyoming Stable Isotope Facility using a Delta Plus XP Continuous Flow Stable Isotope Ratio Mass Spectrometer (Thermo Finnigan, Bremen, Germany) coupled to a Costech Analytical 4010 elemental analyzer. Carbon (13C:12C) and nitrogen (15N:14N) isotope ratios (R) were estimated relative to their respective International Atomic Energy Agency standards (Pee Dee Belemnite limestone and atmospheric nitrogen). Working internal standards were run as controls throughout analysis to ensure accurate measurements. Isotopic values are reported in per mill delta notation \@ref(eq:ratio): -->

@@

For all macroinvertebrate taxa and FFGs, we used least squares linear regression to assess relationships between the baseline’s d15N position at sites along the longitudinal stream gradient.  We wished to compare our baselines’ relationships to the longitudinal gradient with those of the basal resources. So, we did separate least squares linear regressions for two of the basal resources at our sites and compared R2 values between the baselines and the resources at our sites.  The R2 value of our basal resources represent the minimum level of variation that our alternate baselines should explain in order to be used as alternatives to the traditional method (e.g. all basal resources across a season).   
<!-- Baselines with significant ($P \le 0.05$) relationships with the longitudinal gradient and that had R2 values equal to or greater than the basal resources were considered suitable baselines. -->

@@

<!-- \begin{equation} -->

<!-- \end{equation} -->

where TPConsumer is the trophic position of each individual fish at a site, d15NConsumer is the d15N signature of each individual fish, d15NBaseline is the mean d15N signature of the baseline at each site, d15N is the discrimination factor for each trophic transfer, and TLBaseline is the expected trophic level of the baseline. Discrimination factor were estimated using the methods of Caut et al. (2009).

@@

<!-- \begin{equation} -->

<!-- \end{equation} -->

@@

The two basal resources, biofilm and seston, had d15N signatures that were significantly (P > =  0.05) correlated with the longitudinal stream gradient (PC1) (Figure 3; Table 1). All eight of the well distributed taxa and all FFG’s had d15N signatures that were significantly correlated with PC1 (Figure 3). The amount of variation in d15N signatures captured by the longitudinal gradient was similar among the two basal resources, eight sufficiently distributed taxa and the four FFGs.
